---
title: "Pies and Proportions"
output: html_notebook
---

We will look at how to represent basic proportions, emphasising alternatives to Pie charts

Pie Charts
Vertical Stacked Bars Charts
Horizontal Stacked Bars Charts
Pareto Charts
Waffle Charts

## The Pie Chart

I’ll create a pie chart of the composition of the Irish parliament, the Dail, before the election in February 2020.

This data was scraped from the oireachtais web site. I do some preprocessing on it to amalgamate some of the smaller parties/ vacant seats and the ceann comhairle (speaker of the parliament) into an ‘other’ category

```{r}
library(dplyr)
library(igraph)
library(forcats)

dail_32 <- read.csv("ire_dail_parties-2016-20.csv")
dail_32<-dail_32%>%select(c(1,3))


# replace parties with 1 seat with a collective 'other parties'
oth<- dail_32%>%filter(X2020<2)%>%select(X2020)%>%summarise(sum(X2020))
oth$party <-"Other/Vacant"
oth<- oth %>% rename("X2020" ="sum(X2020)" )

# filter our these seats
dail_32<-dail_32%>%filter(X2020>=2)
# add the ' Other' row
colnames(dail_32) <- c("party","X2020")
dail_32<-rbind(dail_32, oth)

# calculate proportions
dail_32$X2020_prop <- round(dail_32$X2020*100/sum(dail_32$X2020),0)

# add a shortened name for each party
dail_32$short <- c("FG", "FF", "SF", "IND", "LAB","S/PBP","SDP", "GP", "OTH" )

# add a display label for each party
dail_32$label <- paste0(dail_32$short,":",dail_32$X2020_prop,"%")

head(dail_32)
```
```{r}
library(scales)

# possible colours
mycols <- c('#a6cee3','#1f78b4','#fb9a99','#b2df8a','#e31a1c','#ff7f00','#fdbf6f','#33a02c','#cab2d6')


party.colours <- c('FG' = '#1f78b4', 'FF' = '#33a02c', 'SF' = 'darkolivegreen', 'LAB' = '#e31a1c',  'GP' = '#b2df8a', 'SDP' ='#cab2d6','#fb9a99', 'IND' = 'darkgrey', 'REN' = '#ff7f00', 'DDI' ='darkgrey','AAA' ='darkgrey', "OTH"='#fb9a99',"S/PBP"= '#fdbf6f')

# a handy way to see colours in your palette. show_col is in the scales package
show_col(party.colours)
```
### Making pie chart

In terms of creating a pie chart, ggplot doesn’t have an oven ready version. Thanks to [this site]((https://www.datanovia.com/en/blog/how-to-create-a-pie-chart-in-r-using-ggplot2/) for insight on how to create it

Compute the position of the text labels as the cumulative sum of the proportion:

Arrange the grouping variable (short) in descending order. This is important to compute the y coordinates of labels. Recall that short refers to the shortened party name I created earlier

To put the labels in the center of pies, we’ll use cumsum(prop) - 0.5*prop as label position.

Key ggplot functions: geom_bar() + coord_polar().

Add text labels: geom_text()

Change fill color manually: scale_color_manual()

Apply theme_void() to remove axes, background, etc


```{r}
library(ggrepel)

set.seed(42)

# sort by 2020 seats a
#dail_32<-dail_32%>%mutate(party = fct_reorder(party, -X2020))%>%arrange(-X2020)

# compute the position of the text labels as the cumulative sum of the proportion:
# Arrange the grouping variable (short) in descending order. This is important to compute the y coordinates of labels.


dail_32<- dail_32 %>%
  arrange(desc(short)) %>%
  mutate(lab.ypos = cumsum(X2020_prop) - 0.5*X2020_prop)


ggplot(dail_32, aes(x = "", y = X2020_prop, fill = short)) +
  geom_bar(width = 1, stat = "identity", color = "white", size=0.1, ) +
  coord_polar("y", start = 0) +
  #geom_text(aes(y = lab.ypos, label = label), color = "white") +
  geom_text_repel(aes(y = lab.ypos, label = label),size=3, segment.color = "darkgrey", segment.size=0.3, nudge_x=0.6) + 
  scale_fill_manual(values = party.colours, name= "Party") +
  theme_void()
```
You can see that the big issue is the difficulty in judging the proportions of the smaller segments.

Also from a practical perspective, the small segments make labelling difficult.

I had to use the ggrepel library for this, which places labels on the plot so that they do not overlap.


