---
title: "Bubble chart over time"
output:
  word_document: default
  html_notebook: default
---

In this section we are going to include all the time data in the data set – so we are going to visualise the GDP and life expectancy trends from 1990 to 2014 for each region in the data set

GDP is a simply a fairly crude economic measure of a country’s wealth. However, it seems to be fact of life that higher GDP in a country entails relatively higher capacity to support the well being of the country’s population.

Our visualisation should uncover that relationship.

I would expect to see some regions to have made relatively short journeys, while others have accelerated to catch up with with the wealthier regions.

### Data Processing using dplyr

I use dplyr functions to produce a view of the nations_data where each region has a ave_gdp_perca (average gdp) and ave_life_expect (average life expectancy).

In order to show the comparative populations per region, the population of each region is calculated by suming the populations of its member countries.

The dplyr library is very useful for processing and filtering data. It is part of tidyverse so you more than likely have it installed

Some of the most useful functions in dplyr:

select Choose which columns to include.
filter Filter the data.
arrange Sort the data, by size for continuous variables, by date, or alphabetically.
group_by Group the data by a categorical variable.
summarize Summarise or aggregate (for each group if following group_by). Often used in conjunction with functions including:
mean Calculate the mean, or average.
median Calculate the median.
max Find| the maximum value.
min Find the minimum value
sum Add all the values together.
n Count the number of records.
mutate Create new column(s) in the data, or change existing column(s).
rename Rename column(s).
bind_rows Merge two data frames into one, combining data from columns with the same name.
These functions can be chained together using the operator %>% which makes the output of one line of code the input for the next. This allows you to run through a series of operations in logical order. I find it helpful to think of %>% as “then.”

### Analysis
As there is high variation in population per region, a simple mean is not appropriate for ave_gdp_perca (average gdp) and ave_life_expect.

As such, I calculate a weighted.mean per region, where the weight is defined by the population proportions of the region’s countries.
```{r}
library(tidyverse)
library(ggplot2 )
library(readr)
library(scales)
    
# load data
nations_data <- read_csv("nations.csv")
```
```{r}
 library(kableExtra)
# summarising the region data. 
# For each region average life expectancy and average gdp per cap is calculated, using a weighted average  - weighted according to the population of the countries in the region

regions<-nations_data %>%
  filter(!is.na(population))%>% # weighted.mean can't handle NAs in the weight vector
  select(region,year, population, gdp_percap, life_expect) %>% 
  group_by(year, region) %>% 
  summarize(totalPop = sum(population,na.rm = TRUE), ave_gdp_percap = weighted.mean(gdp_percap, population, na.rm = TRUE), ave_life_expect = weighted.mean(life_expect, population,na.rm = TRUE ))


  kable(regions, digits = 2, format = "html", row.names = TRUE) %>%
  kable_styling( bootstrap_options = c("striped"),full_width = F,
                font_size = 10) %>%
  
   scroll_box(height = "300px")
```
### Bubble plot

We can now plot this new data using a variation of earlier single plot. The main differences are:

aesthetics are applied to ave_gdp_percap,ave_life_expect and totalPop fields
x and y labels refer to Average GDP per Capita (USD), etc
I added a breaks attribute to the scale_x_log10 function so that more labelled units would show on the x axis (trans_breaks).
The geo_point has a very small alpha value because many of these points are going to overlap. I want the overlap to be clearly perceived.
I’ve explicitly set the alpha value of the colour in the legend to 0.5. Otherwise, it would have the same value as the alpha value of the geom_point - which was too faint


```{r}
p.regions.ave <-
  ggplot(regions,
         aes(
           x = ave_gdp_percap,
           y = ave_life_expect,
           size = totalPop,
           color = region
         )) +
  geom_point(na.rm = TRUE, alpha = 0.25) +
  ggtitle("Trajectory 1990-2014: Average Life Expectancy vs average GDP per Capita per Region") +
  xlab("Average GDP per Capita (USD)") +
  ylab("Average Life Expectancy (years)") +
  # note how scale breaks are inserted
  scale_x_log10(
    labels = trans_format("log10", math_format(10 ^ .x)),
    breaks = scales::trans_breaks("log10", function(x)
      10 ^ x)
  ) +
  scale_color_brewer(palette = 'Dark2', name = "Region") +
  scale_size_area(max_size = 6,
                  labels  = comma,
                  name = "Region \nPopulation") +
  
  
  
  #increase the size of the legend colour points, and their alpha value (too faint otherwise)
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 0.5))) +
  
  
  theme_minimal(base_size = 10) +
  theme(panel.grid.major = element_line(size = 0.2),
        panel.grid.minor = element_line(size = 0.1))

plot(p.regions.ave)
```

###  Labelling the regions
Lets get rid of one of the legends. We will label one data point per region

Let’s choose the data point in each region for the year 1990.

I will use the ggrepel library. Text labels repel away from each other, away from data points, and away from edges of the plotting area.

There is a trick to using ggrepel to make ensure that it doesn’t place labels over unlabelled points : create a column called label. Give every point a blank label value ("") by default and then override this for the points that should actually have a label value.

ggrepel will not create a label for points with a blank label value, but will try to avoid overlapping them, treating as if they were in fact labelled.

```{r}
regions$label <-"" # default is empty string

# my understanding of how ggrepl works is that it will repel labels away from empty labels as well, which is why I set the default labels = ""

# just one label for each region in the year 1990
regions[regions$year==1990,]$label<- regions[regions$year==1990,]$region
```


```{r}
library(ggrepel)

p.regions.ave <-
  ggplot(
    regions,
    aes(
      x = ave_gdp_percap,
      y = ave_life_expect,
      size = totalPop,
      color = region,
      group = region
    )
  ) +
  geom_point(na.rm = TRUE, alpha = 0.25) +
  ggtitle("Trajectory 1990-2014: Average Life Expectancy vs average GDP per Capita per Region") +
  xlab("Average GDP per Capita (USD)") +
  ylab("Average Life Expectancy (years)") +
  # note how scale breaks are inserted
  scale_x_log10(
    labels = trans_format("log10", math_format(10 ^ .x)),
    breaks = scales::trans_breaks("log10", function(x)
      10 ^ x)
  ) +
  scale_color_brewer(palette = 'Dark2', guide = FALSE) +
  scale_size_area(max_size = 6,
                  labels  = comma,
                  name = "Region \nPopulation") +
  
  geom_path(size = 0.2,
            inherit.aes = T,
            colour = "black") +
  #Using geom_label to add text labels
  geom_text_repel(
    aes(label = label),
    size = 3,
    min.segment.length = unit(0, 'lines'),
    segment.size = 0.2,
    segment.color = "dodgerblue2",
    label.padding = unit(0.1, "lines"),
    force = 10,
    label.r = unit(0.15, "lines"),
    colour = "black",
    alpha = 0.5,
    na.rm = TRUE,
    show.legend = FALSE
  )  +
  
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.y = element_line(size = 0.2)) +
  theme(panel.grid.minor.y = element_line(size = 0.1)) +
  theme(panel.grid.major.x = element_line(size = 0.2)) +
  theme(panel.grid.minor.x = element_line(size = 0.1))


plot(p.regions.ave)
```

### Labelling time values

The labelling is mostly fine. However, There are a few problems with this plot

This is a time series plot and there is an implicit sequence of year values presented here - but not an indication of their starting or end values, or in fact the direction. In fact, each point does slightly more to the right from its predecessor - but there’s no reason why GDP and Life Expectancy couldn’t start to reverse.

In short, we need to label the time points - at least the start and end points.

```{r}
# Create a label for the min and max year in each region

regions <- regions %>%
  filter(!is.na(ave_life_expect) & !is.na(ave_gdp_percap)) %>%
  group_by(region) %>% mutate(label = ifelse(year == min(year) |
                                               year == max(year) , year, ""))

p.regions.ave <-
  ggplot(
    regions,
    aes(
      x = ave_gdp_percap,
      y = ave_life_expect,
      size = totalPop,
      color = region,
      group = region
    )
  ) +
  geom_point(na.rm = TRUE, alpha = 0.25) +
  ggtitle("Trajectory 1990-2014: Average Life Expectancy vs average GDP per Capita per Region") +
  xlab("Average GDP per Capita (USD)") +
  ylab("Average Life Expectancy (years)") +
  # note how scale breaks are inserted
  scale_x_log10(
    labels = trans_format("log10", math_format(10 ^ .x)),
    breaks = scales::trans_breaks("log10", function(x) 10 ^ x)) +
 
  scale_color_brewer(palette = 'Dark2', name = NULL) +
  scale_size_area(max_size = 6, labels  = comma, name = "Region Population") +
  
geom_path(size = 0.3, inherit.aes = T, colour="black") +
  
  geom_text_repel(
    aes(label = label),
    size = 2.5,
    min.segment.length = unit(0, 'lines'),
    segment.size = 0.2,
    segment.color = "dodgerblue2",
    colour = "black",
    force = 4,
    alpha = 0.8,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.y = element_line(size=0.2))+
  theme(panel.grid.minor.y = element_line(size=0.1))+
  theme(panel.grid.major.x = element_line(size=0.2))+
  theme(panel.grid.minor.x = element_line(size=0.1))+
  # increase the size of the legend colour points, and their alpha value (too faint otherwise)
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 0.5))) 
  
  plot(p.regions.ave)
```

We’ve replaced the colour legend so as to allow the labelling of year values. I think that this is the prefered solution.

I have added aline through each set of regional points to emphasis the trends.

The geom I used for this is not geom_line which plots a line based on the sequence of x axis values. Instead I am using geom_path which plots a line according to the sequence of the points in the data set – which are ordered by year value ascending.

This is important because there may be situations that you plot where the point ordering is not clear

Here the line helps to emphasis the trend – and we can clearly see the drop in Life expectancy . In Europe in the early nineties – and the thickening of the line in the early 90s in Sub-Sahara Africa suggests the leftward direction towards lower GDP

### A Faceted approach

However, there may be situations where you need to facet the graph. Because each panel represents a region, you no longer need colour to differentiate the series.

When faceting you inevitably lose some of the ease of comparison that a single plot offers. However, the extra space per panel does allow you add extra information that might become too crowded in a single plot.

In the plot below, I’ve calculated the world mean per year for life expectancy and GDP. This acts as a reference line to explain the trends in each series.

For example, you can see that North America in 1990 was ahead of the 2014 world mean for GDP and Life Expectancy.

The GDP in South Asia in 2014 is about that of the 1990 world mean, though life expectancy in South Asia 2014 is higher than the world mean in 1990.

In this section of code, I calculate the world mean per year for life expectancy and GDP. I will plot this line and we will see what regions sit on either side of it

```{r}
nations_mean_per_year<-nations_data %>% 
  filter(!is.na(population))%>% # weighted.can't handle NAS in the weight vector
  select(country, region,year, population, gdp_percap, life_expect) %>% 
  group_by(year) %>% 
  summarize(totalPop = sum(population,na.rm = TRUE), ave_gdp_percap = weighted.mean(gdp_percap, population, na.rm = TRUE), ave_life_expect = weighted.mean(life_expect, population, na.rm = TRUE ))
```

```{r}
library(ggrepel)

  p.regions.ave <-
  ggplot(
    regions,
    aes(
      x = ave_gdp_percap,
      y = ave_life_expect,
      size = totalPop,
      group = region
    )
  ) +
  geom_point(na.rm = TRUE, alpha = 0.15) +
  ggtitle("Trajectory 1990-2014: Average Life Expectancy vs average GDP per Capita per Region") +
  xlab("Average GDP per Capita (USD)") +
  ylab("Average Life Expectancy (years)") +
  # note how scale breaks are inserted
  scale_x_log10(
    labels = trans_format("log10", math_format(10 ^ .x)),
    breaks = scales::trans_breaks("log10", function(x) 10 ^ x)
  ) +
  scale_size_area(max_size = 4, labels  = comma, name = "Region Population") +
  
  geom_path(data=nations_mean_per_year, aes(x = ave_gdp_percap,
      y = ave_life_expect), size = 0.2, colour = "red", inherit.aes = F) +
    geom_path(size = 0.2, inherit.aes = T, colour="black") +
  
  geom_text_repel(
    aes(label = label),
    size = 2.5,
    min.segment.length = unit(0, 'lines'),
    segment.size = 0.15,
    colour = "black",
    segment.color = "dodgerblue2",
    force = 15,
    alpha = 0.8,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.y = element_line(size=0.2))+
  theme(panel.grid.minor.y = element_line(size=0.15))+
  theme(panel.grid.major.x = element_line(size=0.2))+
  theme(panel.grid.minor.x = element_line(size=0.15))+
  theme(axis.title.x = element_text(hjust = -0.01))+
  theme(legend.title = element_text(size = 8))+
  theme(legend.box.background = element_rect(colour = "grey", size=0.2))+
  # increase the size of the legend colour points, and their alpha value (too faint otherwise)
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 0.5)))  +
    
    facet_wrap(vars(region))
```

```{r}
shift_legend2 <- function(p, position) {
  # ...
  # to grob
  gp <- ggplotGrob(p)
  facet.panels <- grep("^panel", gp[["layout"]][["name"]])
  empty.facet.panels <- sapply(facet.panels, function(i) "zeroGrob" %in% class(gp[["grobs"]][[i]]))
  empty.facet.panels <- facet.panels[empty.facet.panels]

  # establish name of empty panels
  empty.facet.panels <- gp[["layout"]][empty.facet.panels, ]
  names <- empty.facet.panels$name
  # example of names:
  #[1] "panel-3-2" "panel-3-3"

# now we just need a simple call to reposition the legend
  reposition_legend(p, position, panel=names)
}
```

```{r}
library(ggplot2)
library(gtable)
library(lemon)
```

```{r}
shift_legend2(p.regions.ave, "center")
```

