---
title: "Fre ploygon and density plots"
output: html_notebook
---

### Data

In the r code below, I create a sythetic data set of height values for men and women based on the following statistics

Male irish mean height 179 cm with SD 7.6 cm
Female irish mean height 164 cm with SD 7.2 cm

```{r}
library(knitr)
library(kableExtra)

set.seed(12566)

height_range <- seq(80, 240, by = 1)

# Male irish  mean height 179 cm with SD  7.6 cm
male_height <- round(rnorm(height_range, mean = 179, sd = 7.6),2)
# Female irish  mean height 164 cm with SD 7.2 cm
female_height <- round(rnorm(height_range, mean = 163, sd = 7.2),2)

m<- factor(c("male","female"))

male <- rep(m[1],length(height_range))
female <- rep(m[2],length(height_range))

m_height<-data.frame(male_height,male)
names(m_height) <- c("height", "gender")
f_height<-data.frame(female_height,female)
names(f_height) <- c("height", "gender")

heights<- rbind(m_height, f_height)


#calculate means for the male and female cohort
library(plyr)
mu <- ddply(heights, "gender", summarise, grp.mean=mean(height))

kable(heights) %>%
  kable_styling( full_width = F)%>%
  scroll_box(height = "300px")
```
### Summarise the data into intervals

```{r}
bin_width <- 5
bin_breaks <- seq(140, 200, bin_width)

height_table<-as.data.frame(table(cut(heights$height, breaks=bin_breaks)))

colnames(height_table) <-c("Height_Range", "Count")

colnames(height_table) <-c("Height_Range", "Count")

kable(height_table) %>%
  kable_styling( full_width = F)
```

The hist function is a more efficient way of summarising values into intervals. I have used the cut function above because it provides the inteval in ( ] form whuhc I could show in a table. However, the output of hist has the same information and more, including mid point of intervals.

```{r}
height_hist2<-hist(heights$height, breaks = bin_breaks, plot = FALSE)
```

### Histograms

The first histogram just summarises the height values in the data set

The bin width is set to 3.

Note how I have added the mean line using the geom_vline function

Try this with a higher bin width and describe the change in terms of how the distribution is summarised

```{r}
bin_width <- 3
bin_breaks <- seq(140, 200, bin_width)


library(ggplot2)
ggplot(heights, aes(x=height)) + 
  
  geom_histogram(binwidth=bin_width, 
                 fill = "steelblue3", col = "white", 
                 position="identity", stat="bin",
                 breaks=bin_breaks, 
                 closed="right") +
  
  scale_x_continuous(expand=c(0,0), breaks = bin_breaks)+
  theme_minimal()+
  theme(legend.position="top") +
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.margin = margin(t = 2, r = 10, b = 2, l = 2, "pt"))
```

### Frequency Polygon

An alternative to the histogram is the frequency polygon, which instead of using bars uses a line that traces the frequency value of each interval. The x=axis value for each interval is its midpoint.

As with the histogram you can control the bin width for the frequency polygon. A larger bin width smooths the polygon.

Frequency polygons are a graphical device for understanding the shapes of distributions. They serve the same purpose as histograms, but are better able to visualise simultaneous plotting of multiple distribution

The plot below of a frequency polygon on top of a histogram is for illustration purposes only.


```{r}
library(ggplot2)
ggplot(heights, aes(x = height)) +
  geom_histogram(
    binwidth = bin_width,
    fill = "grey",
    alpha = 0.5,
    col = "white",
    position = "identity",
    stat = "bin",
    breaks = bin_breaks,
    closed = "right"
  ) +
  
  geom_freqpoly(
    binwidth = bin_width,
    colour = "steelblue3",
    stat = "bin",
    breaks = bin_breaks,
    closed = "right"
  ) +
  
  scale_x_continuous(expand = c(0, 0), breaks = seq(130, 200, by = 5)) +
  theme_minimal() +
  theme(legend.position = "top") +
  theme(
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    plot.margin = margin(
      t = 2,
      r = 10,
      b = 2,
      l = 2,
      "pt"
    )
  )
```
```{r}
library(ggplot2)

ggplot(heights, aes(x=height)) + 
  
 geom_freqpoly(binwidth=bin_width, 
               colour = "steelblue3",
               stat="bin", 
               breaks=bin_breaks, 
               closed="right") +

  scale_x_continuous(expand=c(0,0), breaks = seq(130,200, by=5)) +
  theme_minimal()+
  theme(legend.position="top") +
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```
### Density Plot
```{r}
library(ggplot2)

ggplot(heights, aes(x=height)) + 
  
  geom_density(fill = "steelblue3", 
               col = "steelblue4", 
               bw = 1.5) +
  
  scale_x_continuous(expand=c(0,0), breaks = seq(130,200, by=5)) +
  theme_minimal() +
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```
### Sources:

https://ggplot2.tidyverse.org/reference/geom_boxplot.html https://ggplot2.tidyverse.org/reference/geom_violin.html http://www.sthda.com/english/wiki/ggplot2-violin-plot-quick-start-guide-r-software-and-data- https://www.r-graph-gallery.com/ridgeline-plot/ https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html https://homepage.divms.uiowa.edu/~luke/classes/STAT4580/boxetc.html#ridgeline-plots
